---
title: "CA of also data"
author: "Xu Zhang"
date: "2025-05-27"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## data summary

data summary. Distribution of focus and position of also in HK and GB English.

```{r }
also<-read.csv("also.csv", sep=";", stringsAsFactors=TRUE)

summary(also)

#removing empty rows
cleaned_also <- also[rowSums(is.na(also) | also == "") < ncol(also), ]
cleaned_also$Variety <- droplevels(cleaned_also$Variety)
cleaned_also$Focus <- droplevels(cleaned_also$Focus)
cleaned_also$Position <- droplevels(cleaned_also$Position)
cleaned_also$Relative_Position <- droplevels(cleaned_also$Relative_Position)
cleaned_also$be.also <- droplevels(cleaned_also$be.also)
cleaned_also$R_S_P <- droplevels(cleaned_also$R_S_P)
summary(cleaned_also)

```

## Make a table and ploting the general data distribution

```{r}
#  table for focus and position
library("FactoMineR")
library("factoextra")
library(ISLR)
library(tidyverse)
library(Rfast)
library(MASS)
table(cleaned_also$Variety, cleaned_also$Position, cleaned_also$Focus)
df_table <- as.data.frame(table(cleaned_also$Variety, cleaned_also$Position, cleaned_also$Focus))
colnames(df_table) <- c("Variety", "Position", "Focus", "Frequency")


library(ggplot2)
ggplot(df_table, aes(x = Variety, y = Frequency, fill = Position)) +
  geom_bar(stat = "identity", position = "dodge") +
  facet_wrap(~ Focus) +  # Creates separate plots for each Focus category
  labs(title = "Distribution by Variety, Position, and Focus",
       x = "Variety",
       y = "Frequency") +
  theme_minimal()


focus_position.table <- prop.table(focus_position.table)
addmargins(hair_eye.table)

```
small dataset
unbalanced data distribution
a few independent variabels
-- try with nnet() first and then glmnet() if overfitting or multicolinearity are spotted

##data regularization
```{r}
table(cleaned_also$Position)

```

## multinomial logistic regression model
```{r}
library(nnet)
cleaned_also$NestedVarietyFocus <- interaction(cleaned_also$Variety, cleaned_also$Focus)  # Create nested variable

model_multinom <- multinom(Position ~ Variety + Focus + NestedVarietyFocus, data = cleaned_also)
summary(model_multinom)

```


### testing p-values

```{r}

z_values <- summary(model_multinom)$coefficients / summary(model_multinom)$standard.errors
p_values <- 2 * (1 - pnorm(abs(z_values)))
print(p_values)


```
###displaying results
```{r, echo=FALSE, results='asis'}
install.packages(c("broom", "kableExtra", "knitr"))
library(broom)
library(kableExtra)
library(knitr)# Load package
tidy(model_multinom, conf.int = TRUE, exponentiate = TRUE) %>%
  kable() %>%
  kable_styling("basic", full_width = FALSE)
```

### multinomial model 2 -- only nested results
```{r}
library(nnet)
model_multinom2 <- multinom(Position ~ Variety + Focus, data = cleaned_also)
summary(model_multinom2)
```

# displaying model2
```{r}
z_values <- summary(model_multinom2)$coefficients / summary(model_multinom2)$standard.errors
p_values <- 2 * (1 - pnorm(abs(z_values)))
print(p_values)

```
```{r echo=FALSE, results='asis}
install.packages(c("broom", "kableExtra", "knitr"))
library(broom)
library(kableExtra)
library(knitr)
tidy(model_multinom2, conf.int = TRUE, exponentiate = TRUE) %>%
  kable() %>%
  kable_styling("basic", full_width = FALSE)
```

#checking multicolinearity
```{r}
library(car)
vif(model_multinom2)
```

# refining model by removing non-significant effects
```{r}

model_refined <- multinom(Position ~ Variety, data = cleaned_also)
summary(model_refined)
z_values <- summary(model_refined)$coefficients / summary(model_refined)$standard.errors
p_values <- 2 * (1 - pnorm(abs(z_values)))
print(p_values)
library(broom)
library(kableExtra)
library(knitr)
tidy(model_refined, conf.int = TRUE, exponentiate = TRUE) %>%
  kable() %>%
  kable_styling("basic", full_width = FALSE)

```

```{r}
library(ggeffects)
ggeffect(model_refined, terms = "Variety") %>%
  plot()
```



# refining model with only nested effects
```{r}
model_refined2 <- multinom(Position ~ NestedVarietyFocus, data = cleaned_also)
summary(model_refined2)
z_values <- summary(model_refined2)$coefficients / summary(model_refined2)$standard.errors
p_values <- 2 * (1 - pnorm(abs(z_values)))
print(p_values)
library(broom)
library(kableExtra)
library(knitr)
tidy(model_refined2, conf.int = TRUE, exponentiate = TRUE) %>%
  kable() %>%
  kable_styling("basic", full_width = FALSE)
```
poor model fitting

# refining model with only focus
```{r}
model_refined3 <- multinom(Position ~ Focus, data = cleaned_also)
summary(model_refined3)
z_values <- summary(model_refined2)$coefficients / summary(model_refined3)$standard.errors
p_values <- 2 * (1 - pnorm(abs(z_values)))
print(p_values)
library(broom)
library(kableExtra)
library(knitr)
tidy(model_refined3, conf.int = TRUE, exponentiate = TRUE) %>%
  kable() %>%
  kable_styling("basic", full_width = FALSE)
```

poor model fitting

# trying with glmnet()
```{r}
library(glmnet)

x <- model.matrix(Position ~ Variety, data = cleaned_also)[,-1]  # Convert categorical to matrix
y <- cleaned_also$Position  # Ensure it's a factor


# Lasso regression (alpha = 1)
x <- model.matrix(Position ~ Variety, data = cleaned_also)[,-1]  # Convert categorical predictor to matrix format
y <- as.numeric(cleaned_also$Position)  # Convert factor to numeric codes
levels(cleaned_also$Position)

str(x)  # Check structure of predictors
str(y)  # Check structure of response variable
sum(is.na(x))  # Check for missing values in predictors
sum(is.na(y))  # Check for missing values in response

cv_model <- cv.glmnet(x, y, family = "multinomial", alpha = 1, nfolds = 3)
sum(is.na(cleaned_also))

plot(cv_model)  # Check optimal lambda

best_lambda <- cv_model$lambda.min  # Get best regularization strength

model_lasso <- glmnet(x, y, family = "multinomial", alpha = 1, lambda = best_lambda)
summary(model_lasso)

```

```{r}

```

# plotting effects 
```{r}

```


```{r}


```



